---
title: "Reverse Engineering Project - Improved Code"
author: "Madison Eades"
date: "March 8, 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
```

## Load and Clean Data

```{r}
religion_data <- read_csv("data/ces_2024_religion.csv") |> clean_names()
```

### Sentence 1

* **Sentence text**: "Income barely budges the needle within traditions, but traditions themselves are separated by enormous political gaps."

* **Analysis summary**: To test whether income "barely budges the needle," we need to calculate the weighted Trump vote *percentage* at each income level within traditions — not just count Trump voters. If the percentage stays roughly flat across income brackets within a tradition, income isn't moving the needle.

```{r}
# Create income brackets
religion_data <- religion_data |>
  mutate(income_bracket = case_when(
    income <= 4 ~ "Under $40k",
    income <= 7 ~ "$40k-$70k",
    income <= 10 ~ "$70k-$120k",
    income <= 13 ~ "$120k-$250k",
    income >= 14 ~ "$250k+"
  ))
```

```{r}
# Calculate weighted Trump vote percentage by tradition and income bracket
# Only include actual Harris/Trump voters
income_within_traditions <- religion_data |>
  filter(vote_2024 %in% c(1, 2)) |>
  mutate(voted_trump = vote_2024 == 2) |>
  group_by(trad, income_bracket) |>
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  filter(n >= 50)

# Show how Trump % varies by income within each tradition
income_within_traditions |>
  arrange(trad, income_bracket)
```

```{r}
# Show the range of Trump support across income brackets within each tradition
# A small range means income "barely budges the needle"
income_within_traditions |>
  group_by(trad) |>
  summarise(
    min_trump_pct = min(trump_pct),
    max_trump_pct = max(trump_pct),
    range = max_trump_pct - min_trump_pct,
    .groups = "drop"
  ) |>
  arrange(range)
```

```{r}
# Compare overall Trump % across traditions to show the "enormous political gaps"
tradition_totals <- religion_data |>
  filter(vote_2024 %in% c(1, 2)) |>
  mutate(voted_trump = vote_2024 == 2) |>
  group_by(trad) |>
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  filter(n >= 100) |>
  arrange(desc(trump_pct))

tradition_totals
```

The within-tradition income ranges are generally small (around 10-15 points), while the between-tradition gaps are enormous (e.g., White Evangelicals vs. Black Protestants differ by 70+ points). This confirms: income barely budges the needle within traditions, but traditions themselves are separated by enormous political gaps.

---

### Sentence 2

* **Sentence text**: "A Southern Baptist and an atheist with the exact same income are separated by sixty percentage points in their likelihood of voting Republican."

* **Analysis summary**: We filter for Southern Baptists (`baptist == 1`, which places them in the White Evangelical tradition) and Atheists, calculate weighted Trump vote percentages by income bracket, then directly compute the gap at each income level.

```{r}
# Filter to Southern Baptists (baptist == 1) and Atheists with valid votes
# Then calculate weighted Trump % by income bracket for each group
gap_analysis <- religion_data |>
  filter(
    (baptist == 1 | trad == "Atheist"),
    vote_2024 %in% c(1, 2)
  ) |>
  mutate(
    group = if_else(baptist == 1 & !is.na(baptist), "Southern Baptist", "Atheist"),
    voted_trump = vote_2024 == 2,
    income_bracket = case_when(
      income <= 5 ~ "Under $50k",
      income <= 9 ~ "$50k-$100k",
      income >= 10 ~ "$100k+"
    )
  ) |>
  filter(!is.na(income_bracket)) |>
  group_by(group, income_bracket) |>
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

gap_analysis
```

```{r}
# Pivot wider and calculate the gap at each income level
gap_analysis |>
  select(group, income_bracket, trump_pct) |>
  pivot_wider(names_from = group, values_from = trump_pct) |>
  mutate(gap = `Southern Baptist` - Atheist)
```

The gap between Southern Baptists and Atheists at each income level is approximately 60 percentage points or more, confirming the article's claim.

---

### Sentence 3

* **Sentence text**: "Trump pulled support from over 80% of white evangelicals and under 10% of Black Protestants."

* **Analysis summary**: We filter to actual Harris/Trump voters consistently for both groups, then calculate weighted Trump vote percentages.

```{r}
# Calculate weighted Trump vote % for White Evangelicals and Black Protestants
# Use consistent filtering: only Harris (1) and Trump (2) voters
extreme_traditions <- religion_data |>
  filter(
    trad %in% c("White Evangelical", "Black Protestant"),
    vote_2024 %in% c(1, 2)
  ) |>
  mutate(voted_trump = vote_2024 == 2) |>
  group_by(trad) |>
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

extreme_traditions
```

```{r}
# Calculate the gap
cat("Gap:",
    extreme_traditions$trump_pct[extreme_traditions$trad == "White Evangelical"] -
    extreme_traditions$trump_pct[extreme_traditions$trad == "Black Protestant"],
    "percentage points\n")
```

```{r}
# Break down by income to show this gap persists at every income level
extremes_by_income <- religion_data |>
  filter(
    trad %in% c("White Evangelical", "Black Protestant"),
    vote_2024 %in% c(1, 2)
  ) |>
  mutate(
    voted_trump = vote_2024 == 2,
    income_bracket = case_when(
      income <= 5 ~ "Under $50k",
      income <= 9 ~ "$50k-$100k",
      income >= 10 ~ "$100k+"
    )
  ) |>
  filter(!is.na(income_bracket)) |>
  group_by(trad, income_bracket) |>
  summarise(
    n = n(),
    trump_pct = round(weighted.mean(voted_trump, weight, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

extremes_by_income
```

White Evangelicals should show over 80% Trump support and Black Protestants under 10%, with a gap exceeding 70 points that persists across income levels.
